---
title: "Intubation TMLE"
author: "Tarragona Datathon"
date: '2022-11-11'
output: rmarkdown::pdf_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
```

```{r Load packages and import data}

setwd(dirname(rstudioapi::getSourceEditorContext()$path))

library(tidyverse)
library(lubridate)
library(dplyr)
library(tidyr)
library(ggplot2)
# library(dagitty)
# library(ggdag)
library(data.table)
# install.packages("remotes")
#install.packages("lifecycle")
#install.packages("Rsolnp")
#install.packages("speedglm")
# remotes::install_github("tlverse/tmle3")
# remotes::install_github("tlverse/tmle3mediate")
library(tmle3mediate)
library(tmle3)
library(sl3)
library(speedglm)
library(Rsolnp)
library(lifecycle)


```

# Import Cleaned Data
```{r}
#tinytex::install_tinytex()

# Load data
cleaned_data <- read_csv("Tab1 H3.csv")

```

# Create Variables
Note processing and basic outcomes conducted in Python
```{r Process data}

cleaned_data$sex_male <- as.numeric(cleaned_data$patientsex == 'M')
cleaned_data$outcome <- as.numeric(cleaned_data$hospital_outcome == "EXITUS")

# Create working dataset
ObsData <-subset(cleaned_data, select=c(age,sex_male,bmi,
                                        sofa.max,sofa.avg,
                                        outcome
                                        ))
ObsData <- na.omit(ObsData) 
sapply(ObsData, class)

PatsIDs <-subset(cleaned_data, select=c(age,a_patientid,
                                        sex_male,bmi,
                                        sofa.max,sofa.avg,
                                        outcome
                                        ))
PatsIDs <- na.omit(PatsIDs) 

head(ObsData,n=2)

```

# The Logic of Early Warning Scores 
Early warning scores use patient observations to trigger a review by clinicians if they reach a certain threshold
Appropriate triggering will allow for early review before significant deterioration, allowing for starting of relevant treatment, and hopefully better outcomes

```{r}

# dagify(
#   Obs ~ Patient,
#   Trigger ~ Obs,
#   Review ~ Trigger,
#   Treatment ~ Review,
#   Improve ~ Treatment,
#   Deteriorate ~ Treatment
# ) %>%
#   ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
#   geom_dag_point(size=20) +
#   geom_dag_edges(linemitre=2) +
#   geom_dag_text(size=3)+
#   theme_dag()
# 

```
# Failure to Trigger Review
Differences in Pulse oximter performance has been shown between ethnicities, due to failure in infrared technology to be calibrated apropriately to differnt skin tones

This device is used in Early Warning Scores to and therefore could result in delayed review and treatment
This could result in worse patient outcomes in these groups

```{r}

# dagify(
#   Trigger ~ Obs,
#   Outcome ~ Trigger,
#   Obs ~ Oximeter,
#   Trigger ~ Oximeter
# ) %>%
#   ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
#   geom_dag_point(size=20) +
#   geom_dag_edges(linemitre=2) +
#   geom_dag_text(size=3)+
#   theme_dag()

```

We therefore investigated the effects of pulse oximetry variable performance on different ethncities
Relating this to the TMLE structure this looks like...

```{r}

# dagify(
#   Patient ~ Age + Sex + PMH + Ethnicity,
#   Obs ~ Patient,
#   Outcome ~ Trigger,
#   Trigger ~ Obs,
#   Oximeter ~ Ethnicity,
#   Trigger ~ Obs,
#   Trigger ~ Oximeter
# ) %>%
#   ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
#   geom_dag_point(size=20) +
#   geom_dag_edges(linemitre=2) +
#   geom_dag_text(size=3)+
#   theme_dag()

```


# Prepare the variables 
Y: Primary outcome event (1 outcome, 0 no outcome), this includes any of the following:
  ALIVE
  EXITUS
  
A: ("Treatment") -> SEX

Z: The mediating factor (patients' SOFA max score)

(W: comorbidilities)
W1: Age (Currently floating number)
    Could change to categorical ie over under 65

W2: BMI 

```{r Set variables for TMLE}

# Outcome
Y  <- "outcome"

# Treatment
A  <- "sex_male"

# Mediators
Z = "sofa.max"

# Covariates
W= c("bmi","age")

node_list <- list(
  W = W,
  A = A,
  Z = Z,
  Y = Y
)

```

```{r}
summary(glm(outcome ~ sex_male + sofa.max + bmi + age,ObsData, family=gaussian(link="identity")))
```


# Ensemble Learner
Construct an ensemble learner using a handful of popular machine learning algorithms

```{r}
# SL learners used for continuous data (the nuisance parameter Z)
enet_contin_learner <- Lrnr_glmnet$new(
  alpha = 0.5, family = "gaussian", nfolds = 3
)
lasso_contin_learner <- Lrnr_glmnet$new(
  alpha = 1, family = "gaussian", nfolds = 3
)
fglm_contin_learner <- Lrnr_glm_fast$new(family = gaussian())
mean_learner <- Lrnr_mean$new()
contin_learner_lib <- Stack$new(
  enet_contin_learner, lasso_contin_learner, fglm_contin_learner, mean_learner
)
sl_contin_learner <- Lrnr_sl$new(learners = contin_learner_lib)

# SL learners used for binary data (nuisance parameters G and E in this case)
enet_binary_learner <- Lrnr_glmnet$new(
  alpha = 0.5, family = "binomial", nfolds = 3
)
lasso_binary_learner <- Lrnr_glmnet$new(
  alpha = 1, family = "binomial", nfolds = 3
)
fglm_binary_learner <- Lrnr_glm_fast$new(family = binomial())
binary_learner_lib <- Stack$new(
  enet_binary_learner, lasso_binary_learner, fglm_binary_learner, mean_learner
)
sl_binary_learner <- Lrnr_sl$new(learners = binary_learner_lib)

# create list for treatment and outcome mechanism regressions
learner_list <- list(
  Y = sl_contin_learner,
  A = sl_binary_learner
)

```


# Targeted Estimation of the Natural Indirect Effect

```{r}
require(methods)

tmle_spec_NIE <- tmle_NIE(
  e_learners = Lrnr_cv$new(lasso_binary_learner, full_fit = TRUE),
  psi_Z_learners = Lrnr_cv$new(lasso_contin_learner, full_fit = TRUE),
  max_iter = 1
)
ObsData_NIE <- tmle3(
  tmle_spec_NIE, ObsData, node_list, learner_list
)
ObsData_NIE

```
Based on the output, we see that the indirect effect of the treatment through the mediators (sex male) is 0.008197 - IMV cohort
Based on the output, we see that the indirect effect of the treatment through the mediators (sex male) in all patients H3 is 0.01737

# Targeted Estimation of the Natural Direct Effect

``` {r}
tmle_spec_NDE <- tmle_NDE(
  e_learners = Lrnr_cv$new(lasso_binary_learner, full_fit = TRUE),
  psi_Z_learners = Lrnr_cv$new(lasso_contin_learner, full_fit = TRUE),
  max_iter = 1
)
ObsData_NDE <- tmle3(
  tmle_spec_NDE, ObsData, node_list, learner_list
)
ObsData_NDE

```

From this, we can draw the conclusion that the direct effect of the treatment (through all paths not involving the mediators (ethnicity)) is 0.09847

Together, the estimates of the natural direct and indirect effects approximately recover the average treatment effect, that is, based on these estimates of the NDE and NIE, the ATE is roughly .





## References
https://tlverse.org/tlverse-handbook/causal-mediation-analysis.html
https://migariane.github.io/TMLE.nb.html
